\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-20 NAR_DRAFT.tex
figuring out how to get a web site as a citation

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \texttt{BSgenome}, \texttt{GenomicAlignments}, \texttt{VariantAnnotation}, and \texttt{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification. 
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points}
\label{NAR-ppv_fig}
\end{figure}

>>>>

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \textrm{BSgenome}, \textrm{GenomicAlignments}, \textrm{VariantAnnotation}, and \textrm{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification. 
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=.45\textwidth]{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exomes. 
  Positive predictive value vs. Recall for five different probability cutoffs. 
  Note the restricted axes to improve readability of points}
\label{NAR-ppv_fig}
\end{figure}
